---
title: "Reproducible Research: Peer Assessment 1"
output: 
html_document:
keep_md: true
fig_caption: yes
---

```{r setup, cache=FALSE, include=FALSE}
library(knitr)
opts_chunk$set(fig.width=11, fig.height=5, echo=TRUE)
```


<!-- markdown v1 -->
## Loading and preprocessing 
The first step of this analysis is to load the dataset into the workspace. For the sake of exemplification this dataset is stored in data.raw data.frame:
```{r loading the data}
steps.raw <- read.csv('activity.csv')
steps.raw$date <- as.Date(steps.raw$date,'%Y-%m-%d')
```
To work decently with dates these where converted from strings to valid Date objects.
This document assumes that the data file is in the same folder as this script.

# Data Analyis

To better understand the data structure a simple statistics calculation was performed. The mean value, median total length and valid record length was obtained.

```{r raw data statistics}
#Determines the total number of steps and the number of valid steps
step.raw.stats <-list()
step.raw.stats <- c(step.raw.stats, length(steps.raw$steps))
step.raw.stats <-c(step.raw.stats,sum(as.numeric(is.na(steps.raw$steps))))

#Computes mean and median values
step.raw.stats <- c(step.raw.stats,mean(steps.raw$steps,na.rm = T))
step.raw.stats <-c(step.raw.stats,median(steps.raw$steps,na.rm=T))
names(step.raw.stats) <- c('#total','#valid','mean','median')

kable(as.data.frame(step.raw.stats), format = "markdown")    
```

Some conclusions can be addressed. The dataset is hevelly right skewed being the median much lower than the mean value.
Has seen from the table above there are several missing values in the dataset.
To overcome this issue a reconstruction of the dataset was aimed. 

## Dataset reconstruction
The dataset reconstruction was performed be evaluating the mean and median values for each 5 seconds time interval. 
All the NA values for a given 5 minutes interval were then replaced by the corresponding interval total rounded average along the dataset (excluding Na values).

The first step was to perform  a statistical analysis condition on 5 seconds time intervals.
The code that was used is displayed bellow as the first 10 rows of the resulting data.frame.

```{r raw data conditional statistics}
  #Retrieves for each day the mean and median
 steps.raw.cond.stat <- as.data.frame(sapply(split(steps.raw,steps.raw$interval),function(x) mean(x[,'steps'],na.rm=T)),row.names=NULL)

steps.raw.cond.stat <- cbind(steps.raw.cond.stat,as.data.frame(sapply(split(steps.raw,steps.raw$interval),function(x) median(x[,'steps'],na.rm=T)),row.names=NULL))
colnames(steps.raw.cond.stat) <- c('mean','median')
steps.raw.cond.stat$interval <- as.integer(rownames(steps.raw.cond.stat))
#Displays the first 10 values
kable(head(steps.raw.cond.stat), format = "markdown")  
```



In the second and final step the final dataset was created. 
In this part of the analysis all the NA values in the original dataset were replaced by the rounded average, conditioned on the 5 seconds time interval as shown bellow:
```{r clean dataset creation, results='hide'}
#creates a cleaned dataset
steps.clean <- steps.raw
#Replaces NA values with the mean value


#Cycles throwh the raw data statistico n time interval and assigns the new mwan value
f= function(row){
  steps.clean[steps.clean$interval ==as.numeric(row['interval']) & is.na(steps.clean$steps),'steps'] <<- round(row['mean'])
}
apply(steps.raw.cond.stat, 1, f);
steps.clean.mean <- mean(steps.clean$steps)
steps.clean.total <- sum(steps.clean$steps)
steps.clean.median <- median(steps.clean$steps)
```



# The Analysis

  A shallow descriptive and temporal analysis of the dataset was performed in the following  topics. 
  
## What is mean total number of steps taken per day?

As previously stated the mean number of steps is
```{r total number of steps}
steps.clean.total
```

The mean value of steps is:
```{r Steps mean value}
steps.clean.mean
```

The median value of steps is:
```{r steps median value}
steps.clean.median
```


To determine the number of total steps per day the following code was used.
Performing a histogram analysis on all valid steps yields the following distribution:
```{r Histogram of number of steps}
par(mfrow=c(1,2))
hist(steps.raw$steps, main='Total Steps (raw data)',xlab='Number of steps')
hist(steps.clean$steps, main='Total Steps (cleaned)',xlab='Number of steps')
```

It is concluded that there are no prominent differences in the histograms between the to datasets.
As stated above the mean is way bigger than the median. This is obvious by the huge skewness of the distribution. The histograms also prove an unimodal distribution.
No significant differences on the mean or median are observed. These are backed up by the previously calculated values.

## Conditional step statistics
  For the cleaned dataset it was thought of a conditional statistics on the day of the week. Therefore the following table presents the total number of steps taken for the first 10 days, alongside with the code:
  
```{r per day statistics}
  steps.clean.cond.stat <- as.data.frame(sapply(split(steps.clean,steps.clean$date),function(x) mean(x[,'steps'],na.rm=T)))

steps.clean.cond.stat <- cbind(steps.clean.cond.stat,as.data.frame(sapply(split(steps.clean,steps.clean$date),function(x) median(x[,'steps'],na.rm=T))))
colnames(steps.clean.cond.stat) <- c('mean','median')
                               
kable(head(steps.clean.cond.stat), format = "markdown")     

```


## Step temporal evolution
The evolution of the number of taken steps by day is also an interesting analysis. This can be performed by looking at the number of steps taken each day:

```{r Steps time series}

ts.raw <- as.data.frame(sapply(split(steps.raw,steps.raw$interval),function(x) mean(x[,'steps'],na.rm=T)))

ts.clean<- as.data.frame(sapply(split(steps.clean,steps.clean$interval),function(x) mean(x[,'steps'],na.rm=T)))

#Generates single figure plot for comparissons
plot(rownames(ts.raw),ts.raw[,1],type='l',main='# of Steps time series',xlab='Interval (5m)',ylab='# steps',col='blue')
lines(rownames(ts.clean),ts.clean[,1],type='l',col='red')
legend('topright',c('Raw data','Cleaned data'),col=c('blue','red'), lty=1)
```

The maximum steps were recorded during the following interval:
```{r maximum steps recorded}
steps.clean[steps.clean$steps == max(steps.clean$steps),'interval']
```


## What is the average daily activity pattern?

To determine the average daily activity pattern the comparison between weekdays and weekends was performed.
The first step was to check for every date which were weekends and weekdays. Then two new factors were added to the reconstructed dataset is.weekday and is.weekend


```{r Adding weekday and weekend info to the dataset}
#Stores current colnames in an axiliary vector
aux.colnames <- colnames(steps.clean)

#Retrieves all days that are weekends
steps.clean <- cbind(steps.clean,as.data.frame(sapply(weekdays(steps.raw$date),function(x) any(c('Domingo','Sábado')==x))))

#Determines all days that are weekdays
steps.clean <- cbind(steps.clean,!steps.clean[,4])

#updates colnames
colnames(steps.clean)<- c(aux.colnames,'is.weekend','is.weekday')

#Splits the data base on is weekend or not
aux <- split(steps.clean,steps.clean$is.weekend)
steps.clean.weekend.true <- as.data.frame(aux['TRUE'])
steps.clean.weekend.false <- as.data.frame(aux['FALSE'])

colnames(steps.clean.weekend.true)<- c(aux.colnames,'is.weekend','is.weekday')
colnames(steps.clean.weekend.false)<- c(aux.colnames,'is.weekend','is.weekday')

```

The second step was to plot the data.
```{r Plotting 5 minut time interval for weekend and weekday}

#Averge steps inteverals averaages on weekdays
ts.clean.weekdays <- as.data.frame(sapply(split(steps.clean.weekend.true[,'steps'],steps.clean.weekend.true$interval),function(x) mean(x,na.rm=T)))
#Averge steps inteverals averaages on weekends
ts.clean.weekends <- as.data.frame(sapply(split(steps.clean.weekend.false[,'steps'],steps.clean.weekend.false$interval),function(x) mean(x,na.rm=T)))

#alters colnames
colnames(ts.clean.weekdays) <- c('steps')
colnames(ts.clean.weekends) <- c('steps')

#Plots in same graph for comparisson
plot(rownames(ts.clean.weekends),ts.clean.weekends$steps,type='l',main='# of Steps time series (raw data)',xlab='Interval m',ylab='# steps',col='blue')
lines(rownames(ts.clean.weekdays),ts.clean.weekdays$steps,col='red')
legend('topright',c('Weekends','Weekdays'),col=c('blue','red'), lty=1)
```

It is concluded that on average shows that there is a difference between weekday and weekend activity. The maximum recorded average was on weekends yielding:

```{r Maximum average steps }
max(ts.clean.weekends$steps)
```

